---
title: "Great Tables with **gt**"
subtitle: "How to use **gt**"
date: 2026-03-15
categories: [gt, tables]
description: |
  Experimenting with gt package to learn basics plus
  advanced features.
---

## Goal
The [**{gt}**](https://gt.rstudio.com/) package has so many functions that it can feel overwhelming. This notebook establishes a baseline foundation to create 80% of what you would normally need, and enough conceptual understanding to guess what's possible and where to look.

## Background

There's no best package, just the best for your situation.

```{r}
#| echo: false

dplyr::tribble(
  ~Table                             , ~`Best for`                                           ,
  "`knitr::kble()`/**{kableExtra}**" , "PDF/LaTeX. (Also quick 'n dirty)"                    ,
  "**{flextable}**"                  , "Word/PowerPoint"                                     ,
  "**{gt}**"                         , "HTML reports"                                        ,
  "**{reactable}**"                  , "Interactive HTML dashboards/Shiny apps"              ,
  "**{DT}**"                         , "Quick interactive exploration (Shiny or standalone)" ,
  "What else am I missing?"          , "**{huxtable}**? **{tinytable}**?"
) |>
  gt::gt() |>
  gt::fmt_markdown() |>
  gt::tab_row_group(rows = 6, label = "") |>
  gt::row_group_order(c(NA, ""))
```

## Setup

The package comes with several sample datasets. We'll use `gtcars`, a dataset of high-en sports cars, because it has a mix of numeric and categorical columns.

```{r}
#| code-fold: false

library(tidyverse)
library(gt)

gt_cars_sample <- gt::gtcars |>
  filter(mfr %in% c("Lamborghini", "BMW")) |>
  select(mfr, model, year, hp, mpg_c, mpg_h)

dplyr::glimpse(gt_cars_sample)
```


## Build Up a Table

Let's build up a **{gt}** table from scratch. Here is a reference schematic from the [gt](https://gt.rstudio.com/) website:

![](gt_parts_of_a_table.svg)

The first thing to notice is that the object has a *table body* and *column labels*. It can potentially have a *stub* (row groups or labels), header, and footer.

This is analogous to the **{ggplot}** object. "gt" stands for "grammar of tables", and just like **{ggplot}** layers build from a base, **{gt}** functions pipe together to add structure and styling.

 
### 1. The Basic Table

`gt()` converts a data frame into a `gt_tbl` object. That's all you need, but you can optionally add `tab_header()`, `cols_label()`, `tab_stubhead()`, and `tab_sourcenote()`.

```{r}
#| code-fold: true

tbl_1 <- gt_cars_sample |>
  gt::gt(rowname_col = "mfr") |>
  gt::tab_header(
    title = "Basic Table",
    subtitle = "with stub, column labels, header, and footer."
  ) |>
  gt::cols_label(
    model = "Model",
    year = "Year",
    hp = "Horsepower",
    mpg_c = "MPG - City",
    mpg_h = "MPG - Hwy"
  ) |>
  gt::tab_stubhead("Manufacturer") |>
  gt::tab_source_note("Source: gtcars is from the gt package.")

tbl_1
```


### 2. Footnotes

You can add footnotes almost anywhere on the table. Base command is `tab_footnote(footnote, locations)` where `locations` is a function related to a part of the table:

- `cells_body()`
- `cells_column_labels()`
- `cells_column_spanners()`
- `cells_grand_summary()`
- `cells_group()`
- `cells_row_groups()`
- `cells_source_notes()`
- `cells_stub()`
- `cells_stub_grand_summary()`
- `cells_stub_summary()`
- `cells_stubhead()`
- `cells_title()`

```{r}
#| code-fold: true

tbl_2 <- tbl_1 |>
  gt::tab_footnote(
    "gt Cars is gtcars in gt.",
    locations = gt::cells_title(groups = "title")
  ) |>
  gt::tab_footnote(
    "Horsepower over 600 is a strong vehicle.",
    locations = gt::cells_column_labels(columns = hp)
  ) |>
  # Footnote in table body. `rows` uses a formula.
  gt::tab_footnote(
    "Fuel efficient!",
    locations = gt::cells_body(
      columns = mpg_h,
      rows = mpg_h > 20
    ),
    placement = "right" # puts the marker on the right side of the cell
  ) |>
  gt::tab_header(
    title = "Footnotes",
    subtitle = gt::md(
      "In header (`cells_title()`), <br>column labels (`cells_column_labels()`), and <br>body(`cells_body()`)"
    )
  )

tbl_2
```


### 3. Formatting

`tab_options()` is the **{gt}** equivalent of `ggplot2::theme()`. To format specific cells, use `tab_tyle(style, locations)` where `style` is a function related to a styling type:

- `cell_text()`
- `cell_fill()`
- `cell_borders()`

```{r}
#| code-fold: true

tbl_3 <- tbl_2 |>
  # TABLE DEFAULTS
  gt::tab_options(
    heading.align = "left",
    table.font.size = gt::pct(90)
  ) |>
  # SPECIFIC CELLS
  gt::tab_style(
    style = gt::cell_fill("gray90", alpha = .5),
    locations = gt::cells_column_labels(
      columns = c(model, year, hp, mpg_c, mpg_h)
    )
  ) |>
  gt::tab_header(
    title = "Formatting",
    subtitle = gt::md("with `tab_options()` and `tab_style()`.")
  )

tbl_3
```


### 4. Spanners and Groups

Column spanners group related columns together under a shared label (useful for organizing related metrics). Row groups organize rows by category.

#### Spanners

```{r}
#| code-fold: true

tbl_4 <- tbl_3 |>
  gt::tab_spanner(
    label = "Miles per Gallon",
    columns = mpg_c:mpg_h
  ) |>
  # ALREADY CALLED THIS, BUT YOU CAN OVERRIDE THE PRIOR SETTING
  gt::cols_label(
    mpg_c = "City",
    mpg_h = "Hwy"
  ) |>
  gt::tab_style(
    style = gt::cell_fill("gray95"),
    locations = gt::cells_column_spanners()
  ) |>
  gt::tab_header(
    title = "Column Spanners",
    subtitle = "with `tab_spanner()`"
  )

tbl_4
```

#### Groups

Unlike previous examples, row groups are specified in the base `gt()` call, so we start fresh here.

```{r}
#| code-fold: true

gt_cars_sample |>
  gt::gt(groupname_col = "mfr", rowname_col = "model") |>
  gt::tab_header(
    title = "Row Groups Stacked",
    subtitle = "Groups are separated with headers."
  ) |>
  gt::tab_options(row_group.background.color = "gray95")

gt_cars_sample |>
  gt::gt(
    groupname_col = "mfr",
    rowname_col = "model",
    row_group_as_column = TRUE
  ) |>
  gt::tab_header(
    title = "Row Groups as a Column",
    subtitle = "Groups are inline using `row_group_as_column = TRUE`."
  ) |>
  gt::tab_options(row_group.background.color = "gray95")
```


## Some of What's Possible

At this point, we've covered 80% of what you'd normally need. **{gt}** can do far more than what we've covered, so let's explore some of the possibilities. Here's a quick gallery of advanced techniques. You don't need to master these (at least, I haven't!), but if a use case arises, between the docs and Copilot, you'll figure it out.

::: panel-tabset

### HTML/Markdown

```{r}
#| code-fold: true

tbl_4 |>
  gt::tab_source_note(
    gt::md("**Source**:`gtcars` is from the **{gt}** package.")
  ) |>
  gt::cols_label(hp = gt::html("Hourse<br>Power")) |>
  gt::tab_header(
    title = "Treat Text as Markdown or HTML",
    subtitle = gt::md("with `md()` and `html()`")
  )
```

### Heatmap

```{r}
#| code-fold: true

tbl_4 |>
  gt::data_color(
    columns = hp,
    method = "numeric",
    palette = c("firebrick", "dodgerblue4")
  ) |>
  gt::tab_header(
    title = "Heatmap",
    subtitle = gt::md("with `data_color()`.")
  )
```

### Derived Columns

```{r}
#| code-fold: true

tbl_4 |>
  gt::cols_add(powerful = hp > 600, .after = hp) |>
  gt::cols_add(
    powerful_color = if_else(powerful, "forestgreen", "firebrick")
  ) |>
  gt::cols_hide(powerful_color) |>
  gt::cols_label(powerful = "") |>
  gt::text_case_match(
    "TRUE" ~ fontawesome::fa("dumbbell"),
    "FALSE" ~ fontawesome::fa("frog"),
    .locations = gt::cells_body(columns = powerful)
  ) |>
  gt::tab_style(
    style = gt::cell_text(color = gt::from_column("powerful_color")),
    locations = gt::cells_body(columns = powerful)
  ) |>
  gt::tab_header(
    title = "Derive new cols, hide others",
    subtitle = gt::md("with `cols_add()` and `cols_hide()`")
  )
```

### **{gtExtras}**

```{r}
#| code-fold: true

spark_dat <- gt_cars_sample |> summarize(.by = mfr, spark = list(hp))

gt_cars_sample |>
  inner_join(spark_dat, by = "mfr") |>
  gt::gt() |>
  gtExtras::gt_plt_sparkline(column = spark) |>
  gt::tab_header(
    title = "Add Sparklines (and other plots)",
    subtitle = gt::md("with **{gtExtras}**.")
  )
```

### Interactive Tables

```{r}
#| code-fold: true

tbl_4 |>
  gt::opt_interactive(
    page_size_default = 3
  ) |>
  gt::tab_header(
    title = "Make table Interactive",
    subtitle = gt::md("with `opt_interactive()`")
  )
```
:::

## Takeaways

You now have the mental model and toolkit to create polished tables for 80% of real-world situations. We covered:

- **Structure**: Use stubs, row groups, and column spanners to organize data logically.
- **Context**. Use headers, footers, and footnotes to explain what readers are looking at.
- **Style**. Use colors, text formatting, and heat maps to draw attention to what matters.

From here:

- The [**{gt}** documentation](https::gt.rstudio.com/) is comprehensive and you know the conceptual landscape well enough to navigate it.
- Explore the ecosystem: **{gtSummary}** for statistical summaries, *{gtExtras}** for plots and spark lines.
