---
title: "Understanding **gt** Tables"
subtitle: "How to use **gt**"
date: 2026-03-15
categories: [gt, tables]
description: |
  Experimenting with gt package to learn basics plus
  advanced features.
---

## Goal
The **gt** package has so many functions that it can feel overwhelming. This notebook establishes a baseline foundation to create 80% of what you would normally need, and enough conceptual understanding to guess what's possible and where to look.

## Setup

The package comes with several sample datasets. We'll use `gtcars`.

```{r}
library(tidyverse)
library(gt)

glimpse(gt::gtcars)
```

## Build Up a Table

### 1. The Basic Table

`gt()` converts a data frame into a **gt** object. 

```{r}
#| code-fold: true

tbl_1 <- gt::gtcars |>
    head() |>
    select(mfr, model, year, hp, mpg_c, mpg_h) |>
    # STUB
    gt::gt(rowname_col = "mfr") |>
    # TABLE HEADER
    gt::tab_header(
        title = "gt Cars",
        subtitle = "Like mtcars, but in a different package."
    ) |>
    # COLUMN LABELS
    gt::cols_label(
        # mfr = "Manufacturer",
        model = "Model",
        year = "Year",
        hp = "Horsepower",
        mpg_c = "MPG - City",
        mpg_h = "MPG - Hwy"
    ) |>
    # STUB HEAD
    gt::tab_stubhead("Manufacturer") |>
    # TABLE FOOTER
    gt::tab_source_note("gtcars is from the gt package.")

tbl_1
```

The first thing to notice is that the object has a *table body* and *column labels*. It can potentially have a *stub* (row groups or labels), header, and footer.

![](gt_parts_of_a_table.svg)

This is analagous to the **ggplot** object. There is a base plot with `x` and `y` aesthetic definitions and `labs()`. The analogy is useful because "gt" stands for "grammar of tables" - a loose parallel to "grammar of graphics".[^1]

[^1]: Neither the [gt site](https://gt.rstudio.com/) nor its [GitHub README](https://github.com/rstudio/gt/) mention "grammar of tables", but articles on the web repeatedly attribute the name to it.

::: {.callout-note}
### Grammar of Tables
The *Grammar of Graphics* is a framework of building up a plot with layers. The plot aesthetics map the table attributes to plot attributes. The code does not specify '_draw this here_', but rather 'apply _these rules_ for this geom.'

The *Grammar of Tables* is similar in that the pipeline of **gt** functions applies rules for how to treat the data frame.
:::

### 2. Footnotes

You can add footnotes almost anywhere on the table. Base command is `tab_footnote(footnote, locations)` where `locations` is a function related to a part of the table:

- `cells_body()`
- `cells_column_labels()`
- `cells_column_spanners()`
- `cells_grand_summary()`
- `cells_group()`
- `cells_row_groups()`
- `cells_source_notes()`
- `cells_stub()`
- `cells_stub_grand_summary()`
- `cells_stub_summary()`
- `cells_stubhead()`

```{r}
#| code-fold: true

tbl_2 <- tbl_1 |>
    gt::tab_footnote(
        "Miles per gallon",
        locations = gt::cells_column_labels(columns = mpg_c)
    ) |>
    # Footnote in table body. Uses formula.
    gt::tab_footnote(
        "Fuel efficient!",
        locations = gt::cells_body(
            columns = mpg_h,
            rows = mpg_h > 20
        )
    )

tbl_2
```

`cols_hide()` so you can use the columns without showing them.

## Observations

## Takeaways

- **gtsummary** and **gtextras** extend **gt**.
