---
title: "Virtual Environments"
subtitle: "with **renv**"
date: 2026-04-27
categories: [gt, tables]
description: |
  How to use **renv** for reproducible data science in R.
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
execute:
  echo: true
  warning: false
  error: false
---

Virtual environments ensure that your code runs consistently regardless of where or when it's executed. They allow you to manage project-specific dependencies without affecting your global R or Python installations. 

I've used R for years without virtual environments and nothing has broken. Why should I care about them now?

- For **one-time** analyses, I install any additional packages I need, run my code, and move on. Using **{renv}** feels like overhead with negligible payoff. Only value is if you need to exactly reproduce your results later.
- For **recurring** analyses, I may return to a project and find that some packages need to be installed. Others have been updated, but this is easily resolved. Again, the benefits of **{renv}** don't seem to justify the effort. There is the possibility of subtle drift in results, but that rarely matters. On the other hand, if model defaults change, or factor handling changes, or plot scales or themes change, that can be a real issue.
- For **collaborative** projects, consistent package versions start to make sense. If I upgrade a package to use a new function, that forces my collaborators to upgrade as well.
- For **deployment** to Posit Cloud, builds use whatever package versions are current. If I want to ensure that my app continues to work as I intend, I need to use **{renv}** to lock in the versions.

Generally, the decision framework is use **{renv}** when:

- You need consistency across time (recurring reports)
- Multiple people are involved
- You deploy the project
- You may need to reproduce results exactly

Skip it when:

- Project is exploratory or one-off
- Speed matters more than reproducibility
- You’re the only user and stakes are low


# R: renv (The Standard Approach)

**{renv}** is the recommended tool for managing R package environments. It creates a project-local library and generates a `renv.lock` file that captures exact package versions and sources.

## Basic Workflow

1. **Initialize**: Set up renv for your project
2. **Install Packages**: Add packages as needed during development
3. **Snapshot**: Record current package state to `renv.lock`
4. **Restore**: Recreate the environment from `renv.lock` on another machine

## Core Commands

Initialize renv in your project directory:
```r
# One-time setup per project
renv::init()
```

Install packages during development:
```r
# Install packages as you work
install.packages("dplyr")
install.packages("ggplot2")
```

Snapshot your current environment:
```r
# Record exact versions to renv.lock
renv::snapshot()
```

Restore environment on another machine:
```r
# Install packages exactly as in renv.lock
renv::restore()
```

## How renv Works

- Creates a `renv/` directory with project-specific package installations
- Generates `renv.lock` with exact versions, sources, and dependencies
- Modifies `.Rprofile` to automatically activate the local environment
- Provides tools for managing package conflicts and updates

## Best Practices for R Projects

- Always commit `renv.lock` to version control
- Never commit the `renv/library/` directory (it's large and machine-specific)
- Run `renv::snapshot()` after installing new packages
- Use `renv::status()` to check for discrepancies between lockfile and current environment


# Lockfiles and Reproducibility

Lockfiles are the key to reproducible environments. They record exact versions and sources.

## R: renv.lock

Contains:
- Package names and versions
- Package sources (CRAN, GitHub, local)
- Dependency relationships
- R version information

## Python: requirements.txt vs. Poetry

**requirements.txt** (basic):
```
pandas==2.0.3
numpy==1.24.3
matplotlib==3.7.1
```

**Poetry lockfile** (comprehensive):
- Exact versions and hashes
- Dependency resolution details
- Platform-specific information

## Version Control Strategy

- Commit lockfiles to Git
- Never commit virtual environment directories
- Use CI to validate environment restoration
- Tag releases with specific lockfile versions

# Hands-on Examples

Let's create a simple cross-language data analysis project.

## Project Structure

```
virtual_environs_tutorial/
├── tutorial.qmd
├── R/
│   ├── analysis.R
│   └── shiny_app.R
├── python/
│   ├── app.py
│   └── requirements.txt
├── renv.lock
└── .github/
    └── workflows/
        └── ci.yml
```

## R Analysis Example

Our R script performs basic data exploration and visualization:

```{r}
#| label: r-analysis
#| echo: true
# Load required packages
library(ggplot2)

# Generate sample data
set.seed(123)
df <- data.frame(
  x = rnorm(100),
  y = rnorm(100),
  group = sample(c("A", "B"), 100, replace = TRUE)
)

# Create a plot
p <- ggplot(df, aes(x = x, y = y, color = group)) +
  geom_point() +
  labs(title = "Sample Scatter Plot", x = "X Variable", y = "Y Variable") +
  theme_minimal()

print(p)

# Save the plot
ggsave("R/plot.png", p, width = 6, height = 4)

# Print summary
summary(df)
```

## Shiny App for Deployment

For demonstration purposes, we've included a simple Shiny app:

```{r}
#| label: shiny-app
#| eval: false
#| echo: true
library(shiny)
library(ggplot2)

ui <- fluidPage(
  titlePanel("Virtual Environment Demo"),
  sidebarLayout(
    sidebarPanel(
      sliderInput("n", "Number of points:", 10, 1000, 100),
      selectInput("color", "Color:", c("blue", "red", "green"))
    ),
    mainPanel(plotOutput("plot"))
  )
)

server <- function(input, output) {
  output$plot <- renderPlot({
    set.seed(123)
    df <- data.frame(x = rnorm(input$n), y = rnorm(input$n))
    ggplot(df, aes(x, y)) +
      geom_point(color = input$color) +
      theme_minimal()
  })
}

shinyApp(ui, server)
```

# Deployment to Posit Cloud

Posit Cloud provides an excellent platform for deploying R applications with managed environments.

## Deploying a Shiny App

1. **Prepare Your App**: Ensure all dependencies are in `renv.lock`
2. **Push to GitHub**: Commit your code and lockfiles
3. **Create Posit Cloud Project**: 
   - Go to Posit Cloud
   - Create "New Project from Git Repository"
   - Enter your repository URL
4. **Configure Environment**: Posit Cloud will automatically restore from `renv.lock`
5. **Publish**: Use the publish button to make your app publicly accessible

## Benefits for Deployment

- **Managed Infrastructure**: No server maintenance
- **Automatic Scaling**: Handles traffic spikes
- **Environment Consistency**: Uses your lockfile for exact reproduction
- **Easy Sharing**: Share links with collaborators or for presentations

# GitHub Actions for CI/CD

GitHub Actions can automate environment setup and testing.

## Basic CI Workflow
# yaml
```{text}
name: R + Python CI

on: [push, pull_request]

jobs:
  test:
    runs-on: ubuntu-latest
    strategy:
      matrix:
        r-version: [4.3]
        python-version: [3.10]
    
    steps:
    - uses: actions/checkout@v4
    
    - name: Setup R
      uses: r-lib/actions/setup-r@v2
      with:
        r-version: ${{ matrix.r-version }}
    
    - name: Setup Python
      uses: actions/setup-python@v4
      with:
        python-version: ${{ matrix.python-version }}
    
    - name: Restore R environment
      run: |
        R -e 'install.packages("renv")'
        R -e 'renv::restore()'
    
    - name: Setup Python environment
      run: |
        python -m venv .venv
        source .venv/bin/activate
        pip install -r python/requirements.txt
    
    - name: Run R tests
      run: Rscript R/analysis.R
    
    - name: Run Python tests
      run: |
        source .venv/bin/activate
        python python/app.py
```

## CI Benefits

- **Automated Testing**: Ensures code works in clean environments
- **Environment Validation**: Confirms lockfiles are correct
- **Cross-platform Testing**: Test on multiple R/Python versions
- **Deployment Triggers**: Automatically deploy on successful tests

# Presentation Tips for R Meetup

## Structure Your Talk

1. **Start with Problems**: Show real-world dependency conflicts
2. **Explain Concepts**: Use simple analogies (apartments vs. houses)
3. **Live Demo**: Initialize renv, snapshot, restore on "fresh" machine
4. **Show Cross-language**: Demonstrate reticulate integration
5. **Deployment Demo**: Deploy Shiny app to Posit Cloud
6. **Q&A**: Prepare for common questions

## Key Takeaways for Audience

- Virtual environments are essential for reproducible data science
- renv makes R environment management straightforward
- Python's venv + requirements.txt is the minimum viable approach
- Lockfiles are your insurance policy for deployment
- Start small: use renv on your next R project

## Common Questions

- **"Isn't this overkill for small projects?"** No, the overhead is minimal and prevents future headaches
- **"What about conda?"** Great for complex scientific stacks, but renv/venv are simpler for most R/Python work
- **"How do I handle system dependencies?"** Virtual environments handle packages, but system libraries (like GDAL) need separate management

# Further Resources

- [renv Documentation](https://rstudio.github.io/renv/)
- [Poetry Guide](https://python-poetry.org/docs/)
- [reticulate Package](https://rstudio.github.io/reticulate/)
- [Posit Cloud Deployment](https://docs.posit.co/posit-cloud/)
- [GitHub Actions for R](https://github.com/r-lib/actions)

# Conclusion

Virtual environments transform data science workflows from fragile, machine-dependent processes to robust, reproducible systems. By isolating dependencies and capturing exact versions in lockfiles, you ensure that your analyses will work consistently across different environments and over time. Whether you're an R user dipping into Python or maintaining complex multi-language projects, mastering virtual environments will make you a more effective and confident data scientist.

The examples in this tutorial provide a solid foundation—start by applying renv to your next R project, and you'll quickly see the benefits in practice.